---
title: "Graphics for Beginners"
author: "Matti Meyer"
date: "18.02.2015"
output:
  html_document:
    toc: true
    theme: united

---
#Introduction
This is a R markdown document, to show you some of the basics to develop graphics in R. If you want the data set or more information, look at my [Githubpage](https://github.com/MattiMeyer/R-for-Beginners).

#Packages
```{r, warning=FALSE, message=FALSE}
library(car)
library(knitr)
library(foreign)
library(gplots)
library(psych)
````
```{r, include=FALSE}
opts_chunk$set(warning=FALSE)
```

#Reading the data
We use a data set from the Department of Psychology of the Univerity of Colorado about Interests of people. You can get it also  [here](http://psych.colorado.edu/~carey/Courses/PSYC7291/ClassDataSets.htm). It is a spss data, so we need the package *foreign* to read the data set and to convert it to a data frame.
````{r}
interest <- as.data.frame(read.spss("/home/matti/R-for-Beginners/interest.sav"))
#file.choose() a good help
````

We have to change the label of the gender, because the plots will look better with it:
```````{r}
interest$gender  <- recode(interest$gender, '1="female" ; 2="male" ')

```````

#Graphs for categorial variables

In this chapter we look for nominally and ordinal scaled variables.

##Barplot with absolute frequency

We will create a barplot for the variable "education". Befor that we use the table() function to build a vector:


```{r}
educ<-  table(interest$educ)

barplot(educ)

``````


##Barplot with relative frequencys
Relative frequencys show which percentage have some specific education. Just the y axis ist changing.

``````{r}
barplot(prop.table(educ))
``````

``````{r}
barplot(prop.table(educ), horiz = T ,las=1)
```````

To connect two variables, here education and gender:

``````{r}

barplot(table(interest$educ, interest$gender), legend=T)
`````````



`````{r}
barplot(table(interest$educ, interest$gender), legend=T,beside=T)
```````

##Pie chart
What is about some delicious pie?
`````````{r}
pie(table(interest$educ))
````````

#Graphs for metric variables

We build a barplot with the geometry variable:
````{r}
barplot(interest$geometry)
`````````

Now a histogram:
```````{r}

hist(interest$geometry)

`````

Now a histogram with 10 breaks and a suggested normal distribution:

```````{r}
mean  <-  mean(interest$geometry, na.rm=T)
sd  <-  sd(interest$geometry, na.rm = T)
hist(interest$geometry, breaks=10)
curve(dnorm(x, mean, sd)*length(interest$geometry), add=T)


``````


Sometimes it is difficult to interpretate a histogram, because it depends on the breaks. So we build a density plot with a continous line:

````````{r}

plot(density(interest$geometry, na.rm = T))
curve(dnorm(x, mean, sd), add=T, lty=2)

``````


#Boxplots


Boxplots are important for explorative data analysis and show characteristics of the distribution. Here we use the age:
`````````{r}

boxplot(interest$age)
`````````

You can also create some Boxplots together:
````````{r}
boxplot(data.frame(interest$vocab, interest$reading, interest$sentcomp))
`````````

To show some Boxplots with informations about some groups (e.g. for women and men) you can use the ~ sign. With the *identify()* you can click on the Boxplots, for example to get to know extreme values.

````````{r}
boxplot(interest$age ~  interest$gender)
#identify(interest$gender, interest$age)
```````

#Barplot for means
We want to visualize the mean of education seperated for men and women. To do that we integrate the *tapply()* function in *barplot()*:
```{r}
barplot(tapply(interest$educ, interest$gender, mean, na.rm=T))
```
To integrate more means of variables we also use the tapply function:
```{r}
barplot(tapply(interest$educ, data.frame(interest$gender, interest$age), mean, na.rm=T, na.omit=T), legend=T, xlab="age", ylab="education")
#for easier interpretations add: beside=TRUE
```


#Error bars
To build an error bar, you can use the package gplots and the *plotmeans()* function. This function build confidence intervals and with p=0.95 we set it for the 95% confidence interval:
```{r}

plotmeans(interest$educ~ interest$gender, p=0.95)
```

##Cats-eye plot

If you want to compare the confidence intervalss of some variables use the *error.bars()* function of the psych package.  Here we compare geometry and reading:

```{r}

geo.read <- na.omit(data.frame(interest$geometry, interest$reading))
error.bars(geo.read, within=F,alpha=0.95, eyes=T)
```

To compare means of several groups use the *error.bars.by()* function. But you have to build data.frames:
```{r}
educ <- as.data.frame(na.omit(interest$educ))
gender  <- as.data.frame(na.omit(interest$gender))
error.bars.by(educ, gender, by.var=T, ylim=c(11,13), legend=1, eyes=T, alpha=0.95)
```


##Create own Error bars
To create your own error bars use the *arrow()* function and build arrows around the means of the barplot with a distance of the standard deviation. We do this for the first block in the barplot:
```{r}
des.read  <-  describeBy(interest$reading,interest$gender, mat=T)
plot.read  <- barplot(des.read$mean, names.arg = des.read$group1)
arrows(plot.read[1], des.read$mean[1]-des.read$mean[1], plot.read[1], des.read$mean[1]+des.read$mean[1], code=3, angle=90)

````

To do this for the second or maybe more blocks in the barplot define the i in the following command. It allows you to get all columns in the table of des.read:

````{r}
plot.read  <- barplot(des.read$mean, names.arg = des.read$group1)
for(i in 1:nrow(des.read)){
  arrows(plot.read[i], des.read$mean[i]-des.read$mean[i], 
         plot.read[i], des.read$mean[i]+des.read$mean[i], 
         code=3, angle=90)
}
```

#Q-Q-Plot
When we are interested in a normal distribution histograms can`t help us much. Better are qqplots. The nearer the points are at the line the better is the normal distribution. *qqnorm()* is in the car package:
```{r}
qqnorm(interest$reading)
```
*qqPlot()* is in the car package and shows you the confidence interval. If the point lays near the interval it is normal distributed:

```{r}
qqPlot(interest$reading)
```

#Scatterplot
To get the connection between two variables use scatterplots with two vectors:
```{r}
plot(interest$reading, interest$geometry)
```

##Adding a Lowess-curve
A Lowess curve is a nonparametric curve that rebuilds the connection between two variables, but we need varaibles without missing values:

```{r}
read.geo <- na.omit(data.frame(interest$reading, interest$geometry))
plot(interest$reading, interest$geometry)
lines(lowess(read.geo[,1],read.geo[,2]))
```

##Adding a regression line
To get a regression line use the *abline()* function combinated with a *lm()*. With *identify()* you can again search for extreme values:

```{r}
plot(interest$reading, interest$geometry)
abline(lm(interest$reading~interest$geometry))

```

#The plot() function
The output of the *plot()* function changes with the input. Here a table to visualise it:

plot-function | Diagram
--------|-------
plot(vector)|Index-diagram
plot(factor)| Histogram
plot(vector,vector)| scatterplot
plot(factor, vector)| boxplot
plot(factor, factor)|mosaic-diagram

#Edit graphs
To change the standard setting use the *par()* function to visit the possibilities for changing something:
```{r, eval=FALSE}
par()
help(par)

```

You can add more things in a *plot()* function. Look here:

Adding | Change
------|------------------------------------
pch=x | type of points in a scatterplot
ces=x | size of points or lettering
lty=x | type of line
lwd=x| size of the line
las=1 | numbers on the y axis are shown horizontal
xlab="x", ylab="x"| the names for the axes
xlim=c(min, max), ylim=c(min,max)| the scale of the axes
main="text" | heading
sub="text" | subheading
col=x | the color (*colors()* gives a full list) 

#Adding elements

To add something, write it directly under the *plot()* function, than it will be added to this graph:

Command | Element
-----------|--------------------
abline() | straight line
arrows() | arrows
axis() | lettering axes
curve(....,add=TRUE) | functioncurves
legend() | legend
lines() | lines
points() | points
symbols() | symbols
text() | text

To get more informations use the *help(abline)* function.





